---
title: "Extending the State Space Approach with Mixed Effects"
author: "Thomas P. Vladeck"
date: "`r lubridate::today()`"
output:
  tufte::tufte_html: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tufte)
library(kableExtra)
```


`r newthought("See [this post](https://tomvladeck.com/2020/04/15/rt-ssm.html) first")`. It outlines the bulk of the derivation of this approach, which is just extended to simultaneously model many geographies here. 

```{marginfigure}
If you follow [Kevin Systrom's code](https://github.com/k-sys/covid-19/blob/cde1ee8cee7617992281676b5a144db894ddfecb/Realtime%20R0.ipynb) (this points to a specific commit), it appears he's using a pretty cool Gaussian process approach to model a changing $Rt$ ðŸ‘
```

We are going to leverage the fact that the [Kalman Filter](https://en.wikipedia.org/wiki/Kalman_filter) was built to handle multidimensional observations. So, instead of modeling just one time series, we're going to model _all the time series_ at once. 

The so-called "observation equation" of the Kalman Filter is

$$\theta_t = Z_t a_t$$

In the previous post, we were modeling $\theta_t$ as a single value. Here, we're going to model it as a separate value for every state (along with a nationwide average). 

The $a_t$ are going to be our overall and state-level effects. The trick is going to be in how we set up $Z_t$ so that we can identify the model. 

`r newthought("Let's jump into the code")`. First we get our environment set up:

```{marginfigure}
We need `snakecase` to make the state names easier to work with, and `tictoc` to time some of our code.
```

```{r, message = F, warning = F}
library(tidyverse)
library(KFAS)
library(zoo)
library(snakecase)
library(tictoc)
```


```{r, message = F, warning = F}
url = 'https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv'
dat = read_csv(url)

WINDOW = 20
SERIAL_INTERVAL = 4
GAMMA = 1 / SERIAL_INTERVAL
STATES = dat$state %>% unique
DIM = length(STATES)
```

```{marginfigure}
All this does is to reconstruct the cumulative number of cases within the desired `WINDOW` for each of the states. The code is a bit concise and inscrutable, but it works. 
```

```{r}
dat_multivar = 
  dat %>% 
  filter(state %in% STATES) %>% 
  select(date, state, cases) %>% 
  spread(state, cases) %>% 
  setNames(to_snake_case(colnames(.))) %>% 
  filter(date > lubridate::ymd("2020-03-01")) %>% 
  mutate_at(vars(-date), ~ ifelse(is.na(.x), 0, .x)) %>% 
  mutate_at(vars(-date), function(x) {
    diff(x) %>% 
      {. + 1} %>% 
      {c(rep(0, WINDOW), .)} %>% 
      rollsum(., WINDOW) 
  }) %>% 
  .[-1, ]
```

```{marginfigure}
As before, `itp1` if $I_{t + 1}$ and `it` is $I_t$. The only difference is that these are matrix-valued, not vectors
```

```{r}
itp1 = as.matrix(dat_multivar[-1, 2:ncol(dat_multivar)])
it = as.matrix(dat_multivar[-nrow(dat_multivar), 2:ncol(dat_multivar)])
```

And here is where we construct Z: 

```{r}
observation_matrix = model.matrix(
  ~ 1 + f,
  data = data.frame(f = factor(1:DIM)),
  contrasts = list(f = "contr.sum")
)
```

It's hard to visualize a 55x55 matrix, so let's make a smaller example:

`S this would correspond to the following:

$$\begin{pmatrix}\theta_1 \\ \theta_2 \\ \theta_3 \end{pmatrix} =
  \begin{pmatrix}1 & 1 & 0 \\ 1 &  0 &  1 \\ 1 & -1 & -1 \end{pmatrix} * 
  \begin{pmatrix}a_1 \\ a_2 \\ a_3 \end{pmatrix}$$
  
But it's easier to think of this as   

$$\begin{pmatrix}New\ York \\ Washington \\ Lousiana \end{pmatrix} =
  \begin{pmatrix}1 & 1 & 0 \\ 1 &  0 &  1 \\ 1 & -1 & -1 \end{pmatrix} * 
  \begin{pmatrix}overall\ average \\ state\ effect\ 1 \\ state\ effect\ 2 \end{pmatrix}$$

`r newthought("Thoughts, questions?")`

- [Get at me on twitter](www.twitter.com/tvladeck)
- My email is `tom at gradientmetrics dot com`
- The code for this [is here](https://github.com/tvladeck/covid-rt)

